---
title: "Data Science Homework 2"
author: "ak5357"
date: "2024-09-26"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
```

# Problem 1
### _NYC Transit Data_

**Loading and cleaning the data, keeping only the following variables:**

* Line
* Station Name
* Station Latitude and Longitude
* Routes Served
* Entry
* Vending
* Entrance Type
* ADA Compliance

```{r import_subway_data, message = FALSE}
transit_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", #load data
           na = c("NA", " ", "")) |> #define possible NA values
  janitor::clean_names() |> #standardize column names
  select(line, starts_with("station"), -station_location, #select focus columns
         starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match( #convert entry column from char to lgl
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE,
    )) |> 
  mutate( #ensure all route columns are char type for consistency
    across(
      starts_with("route"), as.character
    ))
```

The dataset contains information related to each entrance and exit for each subway station in New York City.

(Dear Reader: After much consideration, I have come to a difficult decision... Since this question doesn't count for points, I will forego further explanation about the dataset's variables. My sincere apologies. _tips hat_)

The data are not tidy, since there is significant redundance among the columns, specifically those related to the routes.


**Answering questions with data.**
```{r data_questions}
num_distinct_stations = transit_df |> 
  distinct(line, station_name) |> 
  nrow()

num_ada_stations = transit_df |> 
  filter(ada == TRUE) |> 
  nrow()

p_no_vending_yes_entry = transit_df |> 
  summarize(
    proportion = mean(vending == "NO" & entry == TRUE)
    ) |> 
  round(4)
```

Here are some interesting facts about the data:

* The dataset contains information about **`r num_distinct_stations`** distinct stations.
* Of the `r nrow(transit_df)` entrances and exists documented in the dataset, only **`r num_ada_stations`** are ADA compliant.
* The proportion of station entrances / exits without vending that allow entrance is **`r p_no_vending_yes_entry`**, or **`r p_no_vending_yes_entry * 100`**%.

**Reformatting data so that route number and route name are distinct variables.**
```{r reformat_df_by_route}
transit_dr_by_route = transit_df |> 
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route"
  ) |> 
  filter(!is.na(route_name))
```


**Answering more questions with data.**
```{r _distinct_A_stations}
n_distinct_A_stations = transit_dr_by_route |> 
  filter(route_name == "A")  |> 
  distinct(line, station_name) |> 
  nrow()

n_ada_A_stations = transit_dr_by_route |> 
  filter(route_name == "A" & ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
```

Here are some more interesting facts about the data:

* There are **`r n_distinct_A_stations`** distinct stations that serve the A train.
* Of the stations that serve the A train, only **`r n_ada_A_stations`** are ADA compliant.


# Problem 2
### _Mr. Trash Wheel_

**Loading and cleaning the data.**
```{r import_trashwheel_data, eval = FALSE}

read_excel("data/202309 Trash Wheel Collection Data.xlsx",
           sheet = "Mr. Trash Wheel"
           na = c("NA", " ", ""),
           skip = 1
           ) |> 
  janitor::clean_names()
  


```










